---
title: "ABM_orchest"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
date: "2026-04-02"
---

```{r-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10,
                      fig.height = 6)
```

```{r, file_layout, echo = FALSE, }
#[EXISTING] Basic exploration
#[EXISTING] Parameter correlations — minor tweak
#[NEW]      Sensitivity analysis SRC/PCC
#[NEW]      Parameter importance ranking
#[MODIFIED] GA regions — SD params, group percentile,
#insensitive flags, range check, 
#GAML generator
#[EXISTING] Convergence analysis — add violin plot
#[EXISTING] Debate composition — keep
#[EXISTING] Speaking mode — add effect size
#[EXISTING] No change baseline — add limitation flag
#[EXISTING] Best params — add GA bounds check
#[EXISTING] Top 10 — keep
#[EXISTING] Cluster formation — extend
#[EXISTING] Hardest/easiest debates — keep
#[MODIFIED] Failures — justify threshold, 
#add constraint violation flag
#[MODIFIED] Stochasticity — add CV convergence plot
#[EXISTING] Boss table — add VAE if keeping
#[EXISTING] Model comparison — keep
#[EXISTING] Final recommendations — keep
```

```{r, setup_1, echo = FALSE, message = FALSE}
library(tidyverse)
library(dplyr)
library(rvg)
library(broom)
library(officer)
library(flextable)
library(sensitivity)
library(lm.beta)
library(igraph)
library(ggraph)
library(janitor)
library(tidygraph)
library(patchwork)
```

```{r, file-sourcing, include = FALSE, echo = FALSE, cache = FALSE}
source("./functions.R")
source("./data_processing.R")
source("./plots.R")

source("./lhs_analysis_1.R")
#ga_outputs <- source("./ga_analysis.R")$value
#ga_outputs <- source("./ga_analysis.R")$value
# sensitivity, ga, annealing, plots
```

```{r, empirical_comparison, cache = FALSE}
# df comparison test
lhs_outputs$results$comparisons$empirical

# plot of empirical distribution
## needs to add labs/for title and axes and theme
lhs_outputs$plots$empirical_col()

# cross time col plot
lhs_outputs$plots$empirical_cross()

```
# Empirical data discovery
- primary attitude change (8-17% of scale range) occurred between t0-t1
- 3-4% left for simulation during debate (statistically indistinguishable from spontaneous drift in control)



```{r lhs-analysis-comparison, cache = FALSE}
# lhs_analysis comparison - v1 (threshold = 0.001), v2 (threshold = 0.01)

# PCC and PRCC using v2 as df_lhs (combined in lhs_ouputs list)
lhs_outputs$plots$pcc()
lhs_outputs$plots$prcc()

# combined visualization
lhs_outputs$plots$pcc_all()
lhs_outputs$plots$prcc_all()

# include comparison plots and use df_versions
#df_versions <- lhs_outputs$inputs$versions

# interpretation:
# model structure dominates behavior (so threshold is a second-order effect)
# bipolarizaito is the only regime with meaningful sensitivity

# comparison summary
#lhs_outputs$comparisons$summary()

# convergence plot
#lhs_outputs$plots$convergence()


```
# Description of Data in LHS
- v1: 100 fixed cycles 6-5-26
- v2: dynamic conversion 4-5-26


```{r, model-performance, echo = FALSE}
# abm vs ols plots
# summary table, comparison_clean
lhs_outputs$plots$model_performance_rank_main()

# version aware 
lhs_outputs$plots$model_rank_versions()

# comparison with agent heterogeneity (use_distinct_agents)
lhs_outputs$plots$model_performance_uncertainty()

# comparison with gap, how big is the performance gap compared ot best model
lhs_outputs$plots$model_performance_gap()

# tables
## OLS vs ABM comparison (used comparison clean)
lhs_outputs$results$comparisons$ols

## performance ranking
lhs_outputs$results$comparisons$ranking

## Setting up a baseline comparison still results in no_change dominating


```

```{r, behavioral-effects, echo = FALSE}
# Behavioral Effects
## debate composition
lhs_outputs$results$behavioral$composition

## debate composition plot
lhs_outputs$plots$debate_comp_errors()

## speaking mode comparisons
lhs_outputs$results$behavioral$speaking


```

```{r, convergence-dynamics, echo = FALSE}
# Convergence Dynamics
## Convergence analysis
lhs_outputs$results$dynamics$conv_diff

## Convergence violin plot 
lhs_outputs$plots$convergence()

# convergence_comparison
lhs_outputs$plots$convergence_compar()

# opinion variance by versions (v1 fixed, v2 previous iteration with greater threshold)
lhs_outputs$plots$opin_var_versions()

## trade_off plot
# tradeoff raw plot using df_lhs (base data)
lhs_outputs$plots$tradeoff_raw()

# TODO rework convergence_cycle aggregated (uses aggreagted data by model, debate_id, speaking_mode -- aggregated versions of convergence cycle)
lhs_outputs$plots$tradeoff_agg()

```

```{r, robustness, echo = FALSE}
# Robustness
## stochasticitiy
lhs_outputs$results$behavioral$stochastic

##heterogeneity
lhs_outputs$results$behavioral$heterogen

```

```{r, model-comparison, echo = FALSE}
# 1. Global accuracy first (context setter)
df_lhs_directional %>%
  summarize(mean_correct = mean(pct_correct_dir),
            mean_mae = mean(mean_mae),
            mean_baseline = mean(mean_baseline_mae)) %>%
  flextable() %>% autofit()

# 2. Version summary
lhs_outputs$inputs$version_summary

# 3. ABM vs OLS
lhs_outputs$plots$abm_vs_ols()

# 4. Delta scatter
lhs_outputs$plots$delta_direction_sim_vs_empir()

# 5. Directional summary table (top debates)
lhs_outputs$results$comparisons$directional %>%
  mutate(mae_improvement = mean_baseline_mae - mean_mae) %>%
  slice_min(mean_mae, n = 10) %>%
  flextable() %>% autofit()

# 6. Variance by debate type (inline narrative reference only, drop the full table)
# cite: homogeneous mean var 0.006 vs mixed 0.022

```

# interpretation
On average models achieve 52% directional accuracry, while producing higher MAE than no change
- suggests that dynamics introduce noise that slightly exceeds predictive signal at aggregate level
- the global average masks substantial debate level heterogeneity
-- homogeneous debates consistently achieve ~80% directional accuracry -> mechanics are well specified for opinion proximate groups but poorly calibrated for high variance heterogeneous debates

## Overall
1. global performance is modest but above chance
2. homogeneous conditions drive the positive signal
3. pro reduction agents and heterogeneous debates are systematic failues modes



```{r, micro-interaction-dynamics, echo = FALSE}
# interaction of distribution by model type (density of influence score by model type)
lhs_outputs$plots$influence_model_type()

# saturation rate by condition
lhs_outputs$plots$saturation_by_condition()

# wrong direction by anti/pro
lhs_outputs$plots$direction_by_position()

# directional accuracy by current condition, faceted by model type
lhs_outputs$plots$directional_accuracy()

# WIP - influence leaderboard
lhs_outputs$inputs$influence %>%
  slice_max(influence_score, n = 10) %>%
  mutate(efficiency = (influence_score / n_broadcasts) * 10000) %>%
  flextable() %>%
  set_header_labels(efficiency = "efficiency (×10⁻⁴)") %>%
  autofit()

# anti-agent cross position table (new, answers the question)
df_lhs_interactions %>%
  filter(pro_reduction == 0) %>%
  mutate(sender_is_pro = sender_opinion > 0.5) %>%
  group_by(model_type, sender_is_pro) %>%
  summarize(mean_delta = mean(abs(delta)), .groups = "drop") %>%
  flextable() %>% autofit()

```

# interpretation
1. models beat direcitonal baseline ~80% in homogeneous conditions
2. saturation is 6x higher in heterogeneous vs homogeneous
3. pro reduction agenets show systematic wrong-direction bias

# pro interpretation
- pro leaning speakers move anti agents roughly 2x more than anti leaning speakers
- bipolarization shows largest absolute cross position delta: shows highest wrong direction rates (may be moving anti agents more but not always in the right direction)
- This suggests cross-position influence is operative within the bounded confidence framework, though the absolute magnitude remains small — reflecting the attenuating effect of inertia and retention discount mechanics on entrenched positions (Festinger 1957, Nowak & Vallacher 1998).

``` {r, network-analysis, echo = FALSE}
# initial test of network per debate
lhs_outputs$results$networks$per_debate

# aggregate metrics
lhs_outputs$results$networks$aggregate

# graphs
lhs_outputs$results$networks$graphs

```


```{r, debate-classifier, echo = FALSE}
# variance threshold classifier
df_classifier <- df_lhs %>%
  group_by(selected_debate_id, debate_label) %>%
  summarize(
    mean_var = mean(initial_variance),
    mean_mae = mean(mae),
    best_model = model_type[which.min(mae)],
    .groups = "drop"
  ) %>%
  mutate(predicted_type = ifelse(mean_var < 0.01,
                                 "consensus", "bipolarization"))

df_classifier %>%
  mutate(correct = best_model == predicted_type) %>%
  summarize(accuracy = mean(correct))

```

## debate classifier
- LHS results suggest that initial variance plays a significant role in the predictor of model suitability.
- one could use a preliminary classifier for GA runs to check which model would work "best" for a specific debate
- debates with var < 0.01 are predicted to favor consensus, dynamics while high-variance favor bipolarization

-- next stage set-up
- could classify debate by initial variance, then apply GA optimization for that predicted model
- preliminary classifier is too simple, variance is not the only predictor


```{r, optimization-insights, echo = FALSE}
# GA regions, best configs
lhs_outputs$results$regions$bounds
```

```{r, ppt-function-init, echo = FALSE}
#ppt <- read_pptx()
```

