---
title: "ABM_orchest_v1"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: lumen
date: "2026-04-02"
---

```{r-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 15,
                      fig.height = 10)
```

```{r, file_layout, echo = FALSE, }
#[EXISTING] Basic exploration
#[EXISTING] Parameter correlations — minor tweak
#[NEW]      Sensitivity analysis SRC/PCC
#[NEW]      Parameter importance ranking
#[MODIFIED] GA regions — SD params, group percentile,
#insensitive flags, range check, 
#GAML generator
#[EXISTING] Convergence analysis — add violin plot
#[EXISTING] Debate composition — keep
#[EXISTING] Speaking mode — add effect size
#[EXISTING] No change baseline — add limitation flag
#[EXISTING] Best params — add GA bounds check
#[EXISTING] Top 10 — keep
#[EXISTING] Cluster formation — extend
#[EXISTING] Hardest/easiest debates — keep
#[MODIFIED] Failures — justify threshold, 
#add constraint violation flag
#[MODIFIED] Stochasticity — add CV convergence plot
#[EXISTING] Boss table — add VAE if keeping
#[EXISTING] Model comparison — keep
#[EXISTING] Final recommendations — keep
```

```{r, setup_1, echo = FALSE, message = FALSE}
library(tidyverse)
library(rvg)
library(broom)
library(officer)
library(flextable)
library(sensitivity)
library(lm.beta)
library(igraph)
library(ggraph)
library(janitor)
library(tidygraph)
library(patchwork)
library(knitr)
```

```{r, file-sourcing, include = FALSE, echo = FALSE, cache = FALSE}
source("./functions.R")
source("./data_processing.R")
source("./plots.R")
source("framework_analysis.R")
```

```{r, analysis-calls, include = FALSE, echo = FALSE, cache = FALSE}
# empirical data loading 4/6/26
df_empirical <- empirical_prep("./data/data_complete_anonymised.csv")

# LHS pipeline
if (!exists("processed_lhs")) {
  message("Processing raw LHS data files")
  processed_lhs <- process_run(run_configs$lhs_main)
}

if(exists("processed_lhs")) {
  lhs_bundle <- list(
    sim_inputs = processed_lhs,
    df_empirical = df_empirical
  )
  lhs_outputs <- analyze_processed_run(lhs_bundle)
}

# GA pipeline
if (!exists("processed_ga")) {
  message("Processing raw GA data files")
  processed_ga <- process_run(run_configs$ga_main)
}

if(exists("processed_ga")) {
  ga_bundle <- list( # 4/6/26
    sim_inputs = processed_ga,
    df_empirical = df_empirical
  )
  ga_outputs <- analyze_processed_run(ga_bundle)
}
```

# Preface
- this document considers one version of LHS and one of GA (100 fixed cycles)
-- the other with dynamic convergence (convergence_threshold at 0.01) which is designated as **v2**
- Further all of the data displayed below is only considering M (mixed) debates which constitute a mix of pro and anti reduction agents (a future version to be run next week - form 25/5/26 onwards will introduce better segmentation and cross comparison of H and M debates)
- All analyses where data is not labelled in **v1** or **v2** is run with **v1** (which biases the interpretation as each debate is forced to run until it reaches 100 cycles)
- use_distinct_agents: __TRUE__ (each agent has the same base parameter value but has a standard deviation associated with each parameter) / __FALSE__ (all agents in that debate have the same parameter value -> assumes agents are all the same)


```{r, empirical_comparison, cache = FALSE}
# Beta Distribution Check
lhs_outputs$plots$beta_distance_vs_empir()

# Empirical Outcome Distributions
lhs_outputs$plots$empirical_col()
lhs_outputs$plots$empirical_cross()

# cohens d check
lhs_outputs$results$comparisons$empirical_cohen

```

``` {r, ols-compar, echo = FALSE, cache = FALSE}
# display linear comparisons
if(!is.null(lhs_outputs$results$comparisons$ols)) {
  lhs_outputs$results$comparisons$ols %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Debate-Level Benchmark: ABM MAE vs OLS MAE") %>%
    autofit()
}

# baseline overall performance ranking
if(!is.null(lhs_outputs$results$comparisons$ranking)) {
  lhs_outputs$results$comparisons$ranking %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Global (Unoptimized) Model Performance Rankings") %>%
    autofit()
}
```

# Empirical data discovery
```{r lhs-analysis-comparison, cache = FALSE}
# lhs_analysis comparison - v1 (threshold = 0.001), v2 (threshold = 0.01)

# PCC and PRCC using v2 as df_lhs (combined in lhs_ouputs list)
lhs_outputs$plots$pcc()
lhs_outputs$plots$prcc()

# combined visualization if both versions are present
if (!is.null(lhs_outputs$inputs$versions) && nrow(lhs_outputs$inputs$versions) > 0) {
  lhs_outputs$plots$pcc_all()
  lhs_outputs$plots$prcc_all()  
} else {
  message("Single LHS version found, skipping combined PCC/PRCC analysis")
}

```
# PCC and PRCC (Pearson (Partial) Correlation Coefficients)
## Description of Data in LHS
- v1: 100 fixed cycles 6-5-26
- v2: dynamic convergence 4-5-26
- Mean Absolute Error (MAE): unoptimized parameter iterations show weak relationships with raw MAE metrics

```{r, directional-accuracy, echo = FALSE}
# 1. Global accuracy first (context setter)
if (!is.null(lhs_outputs$results$comparisons$directional)) {
  lhs_outputs$results$comparisons$directional %>%
    summarize(mean_correct = mean(pct_correct_dir),
            mean_mae = mean(mean_mae),
            mean_baseline = mean(mean_baseline_mae)) %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Global Directional Accuracy Aggregates") %>%
    autofit()
}

# sign vs magnitude diangostics
if (!is.null(lhs_outputs$results$comparisons$directional)) {
  lhs_outputs$results$comparisons$directional %>%
    select(selected_debate_id, model_type, pct_correct_dir, mean_mae, mean_baseline_mae) %>%
    mutate(
      Diagnostic_Status = case_when(
        pct_correct_dir >= 0.70 & mean_mae <= mean_baseline_mae ~ "Optimized Structure (Correct Sign & Scale)",
        pct_correct_dir >= 0.70 & mean_mae >  mean_baseline_mae ~ "Parameter Attenuation Floor (Correct Sign, Overshot Scale)",
        pct_correct_dir <  0.70 & mean_mae <= mean_baseline_mae ~ "Stochastic Coincidence (Wrong Sign, Low Scale Variance)",
        TRUE
      )
    ) %>%
    group_by(model_type, Diagnostic_status) %>%
    summarize(Debates_count = n(), Average_MAE = mean(mean_mae), .groups = "drop") %>%
    flextable() %>%
    set_caption("Systemic Diagnostic Matrix: Sign Alignment vs Parameter Volatility") %>%
    autofit()
}

# 3. ABM vs OLS
lhs_outputs$plots$abm_vs_ols()

# 4. Delta scatter
lhs_outputs$plots$delta_direction_sim_vs_empir()

# Explicit calling directional script plots
if (!is.null(lhs_outputs$results$comparisons$directional)) {
  # Bar plot for condition accuracy
  print(plot_directional_accuracy(lhs_outputs$results$comparisons$directional))
  
  # four quadrant change scatter plot
  print(plot_delta_direction_scatter(lhs_outputs$results$comparisons$directional))
}

if (!is.null(lhs_outputs$plots$))

```

# Initial Attempt at Variance Debate Classifier
```{r, debate-classifier, echo = FALSE}
# variance threshold classifier
df_classifier <- lhs_outputs$inputs$raw %>%
  group_by(selected_debate_id, debate_label) %>%
  summarize(
    mean_var = mean(initial_variance, na.rm = TRUE),
    mean_mae = mean(mae, na.rm = TRUE),
    best_model = model_type[which.min(mae)],
    .groups = "drop"
  ) %>%
  mutate(predicted_type = ifelse(mean_var < 0.01,
                                 "consensus", "bipolarization"))

df_classifier %>%
  mutate(correct = best_model == predicted_type) %>%
  summarize(accuracy = mean(correct, na.rm = TRUE)) %>%
  mutate(accuracy = round(accuracy, 4)) %>%
  flextable() %>%
  set_caption("Predictive Accuracy of the Initial Variance Classifier") %>%
  autofit()
```
# Debate Classifier Intepretation
- debates initialized below a strict variance ceiling (0.01) match consensus models, wider spreads call for bipolarization configurations (stepping stone for targeted optimization)


# GAML Boundary Layer
- tracks parameter spaces from top 25% performing configurations of LHS runs
``` {r, gaml-boundary-lhs, echo = FALSE, cache = FALSE}
if (!is.null(lhs_outputs$results$regions)) {
  lhs_outputs$results$regions %>%
    select(model_type, use_distinct_agents, best_mae, mae_threshold_used, n, 
           cr_min, cr_max, ct_min, ct_max, rs_min, rs_max, rt_min, rt_max) %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Filtered LHS Base Parameter Envelopes (top 25% slices") %>%
    autofit()
}

```


``` {r, lhs-vs-ga-pcc, echo = FALSE, cache = FALSE}
# initial comparison
if (exists("lhs_outputs") && exists("ga_outputs")) {
  #extrac high leverage parameters based on absolute PRCC values form lhs
  top_sensitivity_params <- lhs_outputs$results$sensitivity$combined$prcc %>%
    filter(output_variable == "mae") %>%
    group_by(model_type) %>%
    slice_max(order_by = abs(value), n = 3, with_ties = FALSE) %>%
    ungroup() %>%
    select(model_type, parameter, prcc_value = value)
  
  # fetch boundaries where GA improved
  ga_realized_regions <- ga_outputs$results$comparisons$ranking
}

#isolate standard deviation envelopes (use_distinct_agents)
# if (!is.null(lhs_outputs$results$regions$regions)) {
#   lhs_outputs$results$regions$regions %>%
#     filter(use_distinct_agents == TRUE) %>%
#     select(model_type, best_mae, cr_min_sd, cr_max_sd, ct_min_sd, ct_max_sd, rs_min_sd, rs_max_sd, rt_min_sd, rt_max_sd) %>%
#     mutate(across(where(is.numeric), ~ round(., 4))) %>%
#     flextable() %>%
#     set_caption("Filtered LHS Micro-Variance Envelopes (SD Parameters)") %>%
#     autofit()
# }

```

``` {r, param-range-audit, echo = FALSE, cache = FALSE}
# check to prevent GA from getting trapped, flags params where width is narrower than 0.05
if (!is.null(lhs_outputs$results$regions$range_check)) {
  if (nrow(lhs_outputs$results$regions$range_check) == 0) {
    cat("### Validation Success: All parameter ranges are sufficient for follow up GA")
  } else {
    lhs_outputs$results$regions$range_check %>%
      select(model_type, use_distinct_agents, parameter, range_ok) %>%
      flextable() %>%
      set_caption("WARNING: param ranges too narrow for GA optimization mobility") %>%
      autofit()
  }
}
```

# Macro comparisons of LHS and GA
``` {r, lhs-vs-ga, echo = FALSE, cache = FALSE}
if (exists("ga_outputs") && !is.null(ga_outputs$results$comparisons$ranking)) {
  p_lhs <- lhs_outputs$plots$model_performance_rank_main() +
    labs(subtitle = "LHS Exploration Footprint (Wide Search")
  
  p_ga <- ga_outputs$plots$model_performance_rank_main() +
    labs(subtitle = "GA Optimization Footprint (Targeted GAML Calibration")
  
  p_lhs + p_ga + plot_layout(guides = "collect")
} else {
  cat("\n\n ** system note: Active GA Package not detected for this specific run",
      "GAML bounds successfully written to disk. run GA engine to append optimization data")
}

```

# Behavioral Spillover & Composition Effects
```{r, behavioral-effects, echo = FALSE}
## debate composition plot
lhs_outputs$plots$debate_comp_errors()

## debate composition
lhs_outputs$results$behavioral$composition %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  flextable() %>%
  set_caption("Debate Composition Behavioral diagnostics") %>%
  autofit()

## speaking mode comparisons
lhs_outputs$results$behavioral$speaking %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  flextable() %>%
  set_caption("Influence of Speaking Mode on Metric Convergence") %>%
  autofit()
```
