---
title: "ABM_orchest_v1"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: lumen
date: "2026-04-02"
---

```{r-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 15,
                      fig.height = 10)
```

```{r, setup_1, echo = FALSE, message = FALSE}
library(tidyverse)
library(rvg)
library(broom)
library(officer)
library(flextable)
library(sensitivity)
library(lm.beta)
library(igraph)
library(ggraph)
library(janitor)
library(tidygraph)
library(patchwork)
library(knitr)
```

```{r, file-sourcing, include = FALSE, echo = FALSE, cache = FALSE}
source("./functions.R")
source("./data_processing.R")
source("./plots.R")
source("framework_analysis.R")
```

```{r, analysis-calls, include = FALSE, echo = FALSE, cache = FALSE}
# empirical data loading 4/6/26
df_empirical <- empirical_prep("./data/data_complete_anonymised.csv")

# LHS pipeline
if (!exists("processed_lhs")) {
  message("Processing raw LHS data files")
  processed_lhs <- process_run(run_configs$lhs_main)
}

if(exists("processed_lhs")) {
  lhs_bundle <- list(
    sim_inputs = processed_lhs,
    df_empirical = df_empirical
  )
  lhs_outputs <- analyze_processed_run(lhs_bundle)
}

# GA pipeline
if (!exists("processed_ga")) {
  message("Processing raw GA data files")
  processed_ga <- process_run(run_configs$ga_main)
}

if(exists("processed_ga")) {
  ga_bundle <- list( # 4/6/26
    sim_inputs = processed_ga,
    df_empirical = df_empirical
  )
  ga_outputs <- analyze_processed_run(ga_bundle)
}
```

# Preface
- this document considers one version of LHS and one of GA (100 fixed cycles)
-- the other with dynamic convergence (convergence_threshold at 0.01) which is designated as **v2**
- Further all of the data displayed below is only considering M (mixed) debates which constitute a mix of pro and anti reduction agents (a future version to be run next week - form 25/5/26 onwards will introduce better segmentation and cross comparison of H and M debates)
- All analyses where data is not labelled in **v1** or **v2** is run with **v1** (which biases the interpretation as each debate is forced to run until it reaches 100 cycles)
- use_distinct_agents: __TRUE__ (each agent has the same base parameter value but has a standard deviation associated with each parameter) / __FALSE__ (all agents in that debate have the same parameter value -> assumes agents are all the same)

# Empirical data discovery
```{r, empirical_comparison, cache = FALSE}
# Beta Distribution Check
lhs_outputs$plots$beta_distance_vs_empir()

# Empirical Outcome Distributions
lhs_outputs$plots$empirical_col()
lhs_outputs$plots$empirical_cross()

# cohens d check
lhs_outputs$results$comparisons$empirical_cohen

```

``` {r, ols-compar, echo = FALSE, cache = FALSE}
# display linear comparisons
if(!is.null(lhs_outputs$results$comparisons$ols)) {
  lhs_outputs$results$comparisons$ols %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Debate-Level Benchmark: ABM MAE vs OLS MAE") %>%
    autofit()
}

# baseline overall performance ranking
if(!is.null(lhs_outputs$results$comparisons$ranking)) {
  lhs_outputs$results$comparisons$ranking %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Global (Unoptimized) Model Performance Rankings") %>%
    autofit()
}
```

```{r lhs-analysis-comparison, cache = FALSE}
# lhs_analysis comparison - v1 (threshold = 0.001), v2 (threshold = 0.01)

# PCC and PRCC using v2 as df_lhs (combined in lhs_ouputs list)
lhs_outputs$plots$pcc()
lhs_outputs$plots$prcc()

# combined visualization if both versions are present
if (!is.null(lhs_outputs$inputs$versions) && nrow(lhs_outputs$inputs$versions) > 0) {
  lhs_outputs$plots$pcc_all()
  lhs_outputs$plots$prcc_all()  
} else {
  message("Single LHS version found, skipping combined PCC/PRCC analysis")
}

```
# PCC and PRCC (Pearson (Partial) Correlation Coefficients)
## Description of Data in LHS
- v1: 100 fixed cycles 6-5-26
- v2: dynamic convergence 4-5-26
- Mean Absolute Error (MAE): unoptimized parameter iterations show weak relationships with raw MAE metrics

```{r, directional-accuracy, echo = FALSE}
# 1. Global accuracy first (context setter)
if (!is.null(lhs_outputs$results$comparisons$directional)) {
  lhs_outputs$results$comparisons$directional %>%
    summarize(mean_correct = mean(pct_correct_dir),
            mean_mae = mean(mean_mae),
            mean_baseline = mean(mean_baseline_mae)) %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Global Directional Accuracy Aggregates") %>%
    autofit()
}

# sign vs magnitude diangostics
if (!is.null(lhs_outputs$results$comparisons$directional)) {
  lhs_outputs$results$comparisons$directional %>%
    select(selected_debate_id, model_type, pct_correct_dir, mean_mae, mean_baseline_mae) %>%
    mutate(
      Diagnostic_Status = case_when(
        pct_correct_dir >= 0.70 & mean_mae <= mean_baseline_mae ~ "Optimized Structure (Correct Sign & Scale)",
        pct_correct_dir >= 0.70 & mean_mae >  mean_baseline_mae ~ "Parameter Attenuation Floor (Correct Sign, Overshot Scale)",
        pct_correct_dir <  0.70 & mean_mae <= mean_baseline_mae ~ "Stochastic Coincidence (Wrong Sign, Low Scale Variance)",
        TRUE ~ "Ambiguous (Wrong Sign, High Scale Variance)"
      )
    ) %>%
    group_by(model_type, Diagnostic_Status) %>%
    summarize(Debates_count = n(), Average_MAE = mean(mean_mae), .groups = "drop") %>%
    flextable() %>%
    set_caption("Systemic Diagnostic Matrix: Sign Alignment vs Parameter Volatility") %>%
    autofit()
}

# 3. ABM vs OLS
lhs_outputs$plots$abm_vs_ols()

# 4. Delta scatter
lhs_outputs$plots$delta_direction_sim_vs_empir()

# Explicit calling directional script plots
if (!is.null(lhs_outputs$results$comparisons$directional)) {
  # Bar plot for condition accuracy
  print(plot_directional_accuracy(lhs_outputs$results$comparisons$directional))
  
  # four quadrant change scatter plot
  print(plot_delta_direction_scatter(lhs_outputs$results$comparisons$directional_agents))
}

if (!is.null(lhs_outputs$plots$direction_by_position)) {
  # called wrapped function from package
  print(lhs_outputs$plots$direction_by_position())
} else if (!is.null(lhs_outputs$inputs$raw)) {
  # fallback to raw batch input
  print(plot_dir_by_pro(lhs_outputs$inputs$raw))
}
```

# Initial Attempt at Variance Debate Classifier
```{r, debate-classifier, echo = FALSE}
# variance threshold classifier
df_classifier <- lhs_outputs$inputs$raw %>%
  group_by(selected_debate_id, debate_label) %>%
  summarize(
    mean_var = mean(initial_variance, na.rm = TRUE),
    mean_mae = mean(mae, na.rm = TRUE),
    best_model = model_type[which.min(mae)],
    .groups = "drop"
  ) %>%
  mutate(predicted_type = ifelse(mean_var < 0.01,
                                 "consensus", "bipolarization"))

df_classifier %>%
  mutate(correct = best_model == predicted_type) %>%
  summarize(accuracy = mean(correct, na.rm = TRUE)) %>%
  mutate(accuracy = round(accuracy, 4)) %>%
  flextable() %>%
  set_caption("Predictive Accuracy of the Initial Variance Classifier") %>%
  autofit()
```
# Debate Classifier Intepretation
- debates initialized below a strict variance ceiling (0.01) match consensus models, wider spreads call for bipolarization configurations (stepping stone for targeted optimization)


# GAML Boundary Layer
- tracks parameter spaces from top 25% performing configurations of LHS runs
``` {r, gaml-boundary-lhs, echo = FALSE, cache = FALSE}
if (!is.null(lhs_outputs$results$regions$bounds)) {
  print(
    lhs_outputs$results$regions$bounds %>%
      select(model_type, use_distinct_agents, best_mae, mae_threshold_used, n, 
             cr_min, cr_max, ct_min, ct_max, rs_min, rs_max, rt_min, rt_max) %>%
      mutate(across(where(is.numeric), ~ round(., 4))) %>%
      flextable() %>%
      set_caption("Filtered LHS Base Parameter Envelopes (top 25% slices") %>%
      autofit()
  )
}

```


``` {r, lhs-vs-ga-pcc, echo = FALSE, cache = FALSE}
# initial comparison
if (exists("lhs_outputs") && exists("ga_outputs")) {
  #extrac high leverage parameters based on absolute PRCC values form lhs
  top_sensitivity_params <- lhs_outputs$results$sensitivity$combined$prcc %>%
    filter(output == "mae") %>%
    group_by(model_type) %>%
    slice_max(order_by = abs(PRCC), n = 3, with_ties = FALSE) %>%
    ungroup() %>%
    select(model_type, parameter, prcc_value = PRCC)
  
  print(top_sensitivity_params)
  
  # fetch boundaries where GA improved
  ga_realized_regions <- ga_outputs$results$comparisons$ranking
}

# isolate standard deviation envelopes (use_distinct_agents)
if (!is.null(lhs_outputs$results$regions$bounds)) {
  
  # Extraction and check if hetero agents exist
  df_sd <- lhs_outputs$results$regions$bounds %>%
    filter(use_distinct_agents == TRUE)
  
  # Only print if there are valid rows
  if (nrow(df_sd) > 0) {
    sd_table <- df_sd %>%
      select(model_type, best_mae, cr_min_sd, cr_max_sd, ct_min_sd, 
             ct_max_sd, rs_min_sd, rs_max_sd, rt_min_sd, rt_max_sd) %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Filtered LHS Micro-Variance Envelopes (SD Parameters)") %>%
    autofit()
    
    print(sd_table)
  } else {
    cat("### Micro-Architecture Note: Current simulation data only uses homogeneous agents")
  }
}

```
# The Evolutionary Bridge
``` {r, param-range-audit, echo = FALSE, cache = FALSE}
# check to prevent GA from getting trapped, flags params where width is narrower than 0.05
if (!is.null(lhs_outputs$results$regions$range_check)) {
  if (nrow(lhs_outputs$results$regions$range_check) == 0) {
    cat("### Validation Success: All parameter ranges are sufficient for follow up GA")
  } else {
    lhs_outputs$results$regions$range_check %>%
      select(model_type, use_distinct_agents, parameter, range_ok) %>%
      flextable() %>%
      set_caption("WARNING: param ranges too narrow for GA optimization mobility") %>%
      autofit()
  }
}
```

# GA results
``` {r, ga-directional-accuracy, echo = FALSE, cache = FALSE}
# 1. Global accuracy first (context setter)
if (!is.null(ga_outputs$results$comparisons$directional)) {
  ga_outputs$results$comparisons$directional %>%
    summarize(mean_correct = mean(pct_correct_dir),
            mean_mae = mean(mean_mae),
            mean_baseline = mean(mean_baseline_mae)) %>%
    mutate(across(where(is.numeric), ~ round(., 4))) %>%
    flextable() %>%
    set_caption("Global Directional Accuracy Aggregates") %>%
    autofit()
}

# sign vs magnitude diangostics
if (!is.null(ga_outputs$results$comparisons$directional)) {
  ga_outputs$results$comparisons$directional %>%
    select(selected_debate_id, model_type, pct_correct_dir, mean_mae, mean_baseline_mae) %>%
    mutate(
      Diagnostic_Status = case_when(
        pct_correct_dir >= 0.70 & mean_mae <= mean_baseline_mae ~ "Optimized Structure (Correct Sign & Scale)",
        pct_correct_dir >= 0.70 & mean_mae >  mean_baseline_mae ~ "Parameter Attenuation Floor (Correct Sign, Overshot Scale)",
        pct_correct_dir <  0.70 & mean_mae <= mean_baseline_mae ~ "Stochastic Coincidence (Wrong Sign, Low Scale Variance)",
        TRUE ~ "Ambiguous (Wrong Sign, High Scale Variance)"
      )
    ) %>%
    group_by(model_type, Diagnostic_Status) %>%
    summarize(Debates_count = n(), Average_MAE = mean(mean_mae), .groups = "drop") %>%
    flextable() %>%
    set_caption("Systemic Diagnostic Matrix: Sign Alignment vs Parameter Volatility") %>%
    autofit()
}

# 3. ABM vs OLS
ga_outputs$plots$abm_vs_ols()

# 4. Delta scatter
ga_outputs$plots$delta_direction_sim_vs_empir()

# Explicit calling directional script plots
if (!is.null(ga_outputs$results$comparisons$directional)) {
  # Bar plot for condition accuracy
  print(plot_directional_accuracy(ga_outputs$results$comparisons$directional))
  
  # four quadrant change scatter plot
  print(plot_delta_direction_scatter(ga_outputs$results$comparisons$directional_agents))
}

if (!is.null(ga_outputs$plots$direction_by_position)) {
  # called wrapped function from package
  print(ga_outputs$plots$direction_by_position())
} else if (!is.null(ga_outputs$inputs$raw)) {
  # fallback to raw batch input
  print(plot_dir_by_pro(ga_outputs$inputs$raw))
}
```

# Macro comparisons of LHS and GA
``` {r, lhs-vs-ga-mae, echo = FALSE, cache = FALSE, fig.width = 16, fig.height = 10}
if (exists("ga_outputs") && !is.null(ga_outputs$results$comparisons$ranking)) {
  p_lhs <- lhs_outputs$plots$model_performance_rank_main() +
    labs(subtitle = "LHS Exploration Footprint (Wide Search")
  
  p_ga <- ga_outputs$plots$model_performance_rank_main() +
    labs(subtitle = "GA Optimization Footprint (Targeted GAML Calibration")
  
  p_lhs + p_ga + plot_layout(guides = "collect")
} else {
  cat("\n\n ** system note: Active GA Package not detected for this specific run",
      "GAML bounds successfully written to disk. run GA engine to append optimization data")
}

# Welch's test comparison for improvement
# extract mae values
lhs_vector <- lhs_outputs$inputs$raw$mae
ga_vector <- ga_outputs$inputs$raw$mae

# run welch's t test
defensive_ttest <- t.test(ga_vector, lhs_vector, paired = FALSE, alternative = "two.sided")

# print results
print(defensive_ttest)

```

``` {r, lhs-vs-ga-convergence, echo = FALSE, cache = FALSE}
# Convergence Comparison (depends on fixed or dynamic convergence)
p_lhs_conv <- lhs_outputs$plots$convergence() + labs(subtitle = "LHS")
p_ga_conv <- ga_outputs$plots$convergence() + labs(subtitle = "GA")
p_lhs_conv + p_ga_conv + plot_layout(guides = "collect")
```

``` {r, lhs-vs-ga-directional, echo = FALSE, cache = FALSE}
# Directional Accuracy Comparison 
combined_directional <- bind_rows(
  lhs_outputs$results$comparisons$directional %>% mutate(run_type = "LHS"),
  ga_outputs$results$comparisons$directional %>% mutate(run_type = "GA")
)

combined_directional %>%
  group_by(run_type, model_type) %>%
  summarize(mean_correct = mean(pct_correct_dir),
            mean_mae = mean(mean_mae), .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  flextable() %>%
  set_caption("LHS vs GA Directional Accuracy Comparison") %>%
  autofit()

```


# Behavioral Spillover & Composition Effects
```{r, behavioral-effects, echo = FALSE}
## debate composition plot
lhs_outputs$plots$debate_comp_errors()

## debate composition
lhs_outputs$results$behavioral$composition %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  flextable() %>%
  set_caption("Debate Composition Behavioral diagnostics") %>%
  autofit()

## speaking mode comparisons
lhs_outputs$results$behavioral$speaking %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  flextable() %>%
  set_caption("Influence of Speaking Mode on Metric Convergence") %>%
  autofit()
```

# Summary & Key Insights
## 1. Statistical Validation of algorithmic progress
A Welch's t-test run on the raw output data, found that the transition from LHS to GA proved statistically significant (<0.01). The GA drove down the MAE from an LHS baseline of $\mu = 0.0875$ to an optimized mean of $\mu = 0.0644$ across all debates ($t(63394) = -48.96, p < 2.2 \times 10^{-16}$). This decrease of roughly 26% in mae demonstrates that the GA is effectively optimizing parameters and we are making progress towards reaching the no_change baseline.

The case for agent heterogeneity: Across all three regimes, models that assume homogeneous agents ("use_distinct_agents" = FALSE, where all agents have the same parameter values without SDs) perform worse than when they have an SD associated to each parameter. More specifically, under the clustering and consensus regimes, incorporating the SDs (distinct agents) dropped the 

## 2. Key Parameter Drivers & Sensitivity (PRCC)
**Consensus** is mostly governed by the parameter "convergence_rate" (speed at which agents making "opinion steps") which reduces mae.
**Clustering & Bipolarization** are driven by the parameter "confidence_threshold" (threshold under which agents interact with others in the same opinion, i.e. "attract") and "repulsion_threshold" (threshold above which agents repulse/refuse to interact with others). These two parameters confirm that the models are working correctly whereby agents create isolated echo chambers of opinion or in severe cases (i.e. bipolarization) two opposite opinion groups.

## 3. The Best Debate Model Fit
The lowest MAE found across all debates and both algorithms is **$\text{MAE} = 0.0081$** which was captured under **Clustering** and **Consensus** models. This was found with heterogeneous agents on debate M24. 

## 4. Interpretation of Sign vs Scale
Evaluating the simulated outputs relies on distinguishing between **directional sign** (are model shifts going in the right direction, wrt T2) and **systemic scale** (the magnitude of this shift). 
**Bipolarization:** the GA successfully matched both sign and scale ($\text{MAE} \approx 0.0425$) implying that the simulation successfully mirrored the real world dynamics of echo chambers.
**Consensus:** This model moved in the correct direction across debates but overshot the empirical scale ($\text{MAE} \approx 0.0352$). 

## 4. Initial Beta Distributions
The wide range of simulated betas (compared with the empirical beta) demonstrates that the ABM parameters are highly sensitive and explory structural spaces rather than outputting a flat random distribution. The fact that the dashed line (empirical beta) cuts through some of the distributions demonstrates that the ABM is not just simulating noise (besides the very weak empirical signal). 
**Bipolarization:** The empirical line cuts right through the center of the simulated distribution, so the model matches the direction and real-world scale of the echo chambers (MAE ~ 0.0425).
**Consensus:** The empirical line intersects the outer tail, implying that the model got the direction right but overshot the scale, simulated consensus was too strong compared with real world groups that have stubborn outliers (MAE ~ 0.0352).
**Clustering:** The empirical line intersects the far right tail of the bimodal simulated distribution. The model only hits the human signal in extreme edges cases, implying that the parameters are on a critical tipping point. The current parameter bounds are too broad, which allows the model to drift away from the human signal.